A Horrocks-type theorem for even orthogonal $K_2$

### 2019-09-20 dl-paper.tex
Minor

After the original Serre problem had been solved, numerous analogous questions drew the attention of specialists (see e.\,g.~\cite{Su77, Su82, Abe83, Tu83, Lam10, St-poly, St-Ded}). For example, A.~Suslin formulated and solved the so-called $K_1$-analogue of Serre's problem. This result asserts that the functor $\K_{1,n}(R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has the property $\K_{1,n}(k[x_1, \ldots x_n]) = \K_{1,n}(k) = k^\times$ for all fields $k$ and $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently, A.~Stavrova has obtained probably the most general results in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on arbitary isotropic reductive group scheme $G$ of isotropic rank $\geq 2$ (see~\cite[Theorem~1.2]{St-poly}).
See also \cite[Corollary~1.2]{St-Ded} for the generalization of Abe's result to Dedekind domains of arithmetic type.

@@

\section{Elementary calculations in relative Steinberg groups}
Throughout this section $\Phi$ denotes a simply laced root system, $R$ a commutative ring, and $I, J$ are two ideals of $R$. 
We denote by $\overline{\St}(\Phi, R, I)$ the kernel of the map $\St(\Phi, R) \to \St(\Phi, R/I)$.
This group coincides with the image in $\St(\Phi, R)$ of the relative group $\St(\Phi, R, I)$ defined in~\cref{sec:quillen}.

@@

Let us mention an immediate application of the just proved lemma. 
First of all, recall the following two results which give two generating sets for the group $\overline{\St}(\Phi, R, I)$.
\begin{externaltheorem}[Stein--Tits--Vaserstein] \label{thm:Tits} For arbitrary $\Phi$ of rank $\geq 2$ the group $\overline{\St}(\Phi, R, I)$ is generated by elements $z_\alpha(s, \xi)$, $\alpha \in \Phi$, $s \in I$, $\xi \in R$. \end{externaltheorem} \begin{proof} See e.\,g.~\cite[Theorem 2]{Va86}. \end{proof}

@@

\begin{externaltheorem}[Stepanov] \label{thm:Stepanov} For arbitrary $\Phi$ of rank $\geq 2$ the group $\overline{\St}(\Phi, R, I)$ is generated as a group by the subset $\mathcal{Z}(\Sigma_S, R, I)$.
 Here $\Sigma_S$ denotes the special part of an arbitrary parabolic subset of roots $S \subseteq \Phi$.
\end{externaltheorem} \begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

\begin{rem} For a simply-laced $\Phi$~\cref{thm:Stepanov} can be deduced from~\cref{thm:Tits} by means of~\cref{Zrels}. Indeed, consider the operator $d \colon 2^\Phi \to 2^\Phi$ of root subsets given by $d(U) = U \cup (U - U)\cap \Phi$, $U \subseteq \Phi$. In other words, $d$ adjoins to $U$ all differences of roots from $U$ which are themselves roots. It is not hard to show that for any parabolic subset $S \subseteq \Phi$ the subset $\Sigma_S$ has the property that $d^n(\Sigma_S) = \Phi$ for some $n>1$ (in fact, $n=2$). It remains to see that relation~\eqref{Z5} immediately implies that every group $G$ containing $\mathcal{Z}(U, R, I)$ also contains $\mathcal{Z}(dU, R, I)$. \end{rem}

@@

\begin{lemma} \label{lem:Tulenbaev-formula} 
One has $S_\alpha(a, x_{-\alpha}(mX)) = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am).$
\end{lemma}
\begin{proof} Since $\Phi$ is nonsymplectic, we can choose $\gamma \in \Phi$ such that $\langle \alpha, \gamma \rangle = -1$. Direct computation using~\eqref{eq:conj-h-x},\eqref{eq:conj-h-h},\eqref{eq:steinberg},\eqref{eq:dennis-stein},\eqref{eq:symbol-properties} shows that
 \begin{multline*}
 x_\alpha(aX^{-1}) \cdot x_{-\alpha}(mX) = {}^{h_\gamma(X)}(x_\alpha(a) \cdot x_{-\alpha}(m)) = \\
 = {}^{h_\gamma(X)}\left( x_{-\alpha}\left(\tfrac{m}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_\alpha\left(\tfrac{a}{1+am}\right) \right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(X^{-1}(1+am))\cdot h_\alpha^{-1}\left(X^{-1}\right) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot \{X^{-1}, 1+am\} \cdot h_\alpha(1+am)\cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) \cdot \{1+am, X\}, \end{multline*}
which implies the assertion. \end{proof}

>>>>

After the original Serre problem had been solved, numerous analogous questions drew the attention of specialists (see e.\,g.~\cite{Su77, Su82, Abe83, Tu83, Lam10, St-poly, St-Ded}). For example, A.~Suslin formulated and solved the so-called $K_1$-analogue of Serre's problem. This result asserts that the functor $\K_{1,n}(R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has the property $\K_{1,n}(k[x_1, \ldots x_n]) = \K_{1,n}(k) = k^\times$ for all fields $k$ and $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently A.~Stavrova has obtained probably the most general results in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on arbitary isotropic reductive group scheme $G$ of isotropic rank $\geq 2$ over a field (see~\cite[Theorem~1.2]{St-poly}) and also generalized Abe's result to Dedekind domains of arithmetic type (see~\cite[Corollary~1.2]{St-Ded}).

@@

\section{Elementary calculations in relative Steinberg groups}
Throughout this section $\Phi$ denotes an irreducible root system of rank $\geq 2$, $R$ a commutative ring, and $I, J$ denote a pair of ideals of $R$.
Unless stated otherwise we assume $\Phi$ to be simply laced.
We denote by $\overline{\St}(\Phi, R, I)$ the kernel of the map $\St(\Phi, R) \to \St(\Phi, R/I)$.
This group coincides with the image in $\St(\Phi, R)$ of the relative group $\St(\Phi, R, I)$ defined in~\cref{sec:quillen}.

@@

@@

Let us mention an immediate application of the just proved lemma.
First of all, recall the following two results which deliver two different generating sets for the group $\overline{\St}(\Phi, R, I)$ (notice that both results apply even if $\Phi$ is not simply laced)
\begin{externaltheorem}[Stein--Tits--Vaserstein] \label{thm:Tits} The group $\overline{\St}(\Phi, R, I)$ is generated (as an abstract group) by elements $z_\alpha(s, \xi)$, $\alpha \in \Phi$, $s \in I$, $\xi \in R$. \end{externaltheorem} \begin{proof} See e.\,g.~\cite[Theorem 2]{Va86}. \end{proof}

Recall that a closed subset $S \subseteq \Phi$ is called {\it parabolic} if $S \cup -S = \Phi$.
Its {\it special part} $\Sigma_S$, by definition, consists of all $\alpha \in S$ such that $-\alpha \not\in S$.
\begin{externaltheorem}[Stepanov] \label{thm:Stepanov} 
Let $S \subseteq \Phi$ be a parabolic subset of $\Phi$. Then the group $\overline{\St}(\Phi, R, I)$ is generated by the set $\mathcal{Z}(\Sigma_S, R, I)$.
 \end{externaltheorem} \begin{proof} See~\cite[Lemma~4]{S15}. \end{proof}

\begin{rem} We claim that in the simply-laced case \cref{thm:Stepanov} can be deduced from~\cref{thm:Tits} by means of~\cref{Zrels}. Indeed, consider the operator $d \colon 2^\Phi \to 2^\Phi$ of root subsets given by $d(U) = U \cup (U - U)\cap \Phi$, $U \subseteq \Phi$. In other words, $d$ adjoins to $U$ all differences of roots from $U$ which are themselves roots. It is not hard to show that for any parabolic subset $S \subseteq \Phi$ the subset $\Sigma_S$ has the property that $d^n(\Sigma_S) = \Phi$ for some $n>1$ (in fact, $n=2$). It remains to see that relation~\eqref{Z5} immediately implies that every group $G$ containing $\mathcal{Z}(U, R, I)$ also contains $\mathcal{Z}(dU, R, I)$. \end{rem}

@@

\begin{lemma} \label{lem:Tulenbaev-formula} 
One has $S_\alpha(a, x_{-\alpha}(mX)) = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am).$
\end{lemma}
\begin{proof} Since $\Phi$ is nonsymplectic, we can choose $\gamma \in \Phi$ such that $\langle \alpha, \gamma \rangle = -1$. Direct computation using~\eqref{eq:conj-h-x}--\eqref{eq:dennis-stein},\eqref{eq:symbol-properties} shows that
 \begin{multline*}
 x_\alpha(aX^{-1}) \cdot x_{-\alpha}(mX) = {}^{h_\gamma(X)}(x_\alpha(a) \cdot x_{-\alpha}(m)) = \\
 = {}^{h_\gamma(X)}\left( x_{-\alpha}\left(\tfrac{m}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_\alpha\left(\tfrac{a}{1+am}\right) \right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(X^{-1}(1+am))\cdot h_\alpha^{-1}\left(X^{-1}\right) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot \{X^{-1}, 1+am\} \cdot h_\alpha(1+am)\cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) = \\
 = x_{-\alpha}\left(\tfrac{mX}{1+am}\right) \cdot \langle a, m\rangle_\alpha \cdot h_\alpha(1+am) \cdot x_{\alpha}\left(\tfrac{aX^{-1}}{1+am}\right) \cdot \{1+am, X\}, \end{multline*}
which implies the assertion. \end{proof}
